VariantSpark: Population Scale Clustering of Genotype Information

### 2015-09-16 doc/publications/journal2014/paper.tex
admixture implementation

We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters. 
For this comparison, we use the \ARI{} (ARI) metric, which returns a value between 1 (perfect) and 0 (random clustering). 
Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
Including AMR individuals, we observe that the majority of these individuals are placed in the same group as Europeans, likely accurately reflecting their migrational backgrounds. 
Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Hunter2014}.

@@

\subsection*{Personal Genome Project}
To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP). The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features ~\cite{Lunshof2013}.
The VCF file we use contains 985,790 variants.

Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
The observed ARI is negative, indicating the clustering is completely random. 
This poor result may be due to the labels being too broad and not accurately reflecting genetic diversity. For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, and Bulgaria, amongst others. 
The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
% TODO provide mapping
We therefore only include individuals where all grandparents were reported to be from the same country leaving 36 individuals from 23 different countries, which we group into their approximate super-populations (see Supplemental Table X). 
Clustering these individuals results in an ARI of 0.29.

Although we see an increase in cluster-accuracy when removing individuals with a mixed background and operating at super-population level, the clustering is not as precise as on the carefully curated and characterised individuals from the 1000 Genomes Project.
%TODO add numbers
This is especially the case since XX\% are 'White' Americans, which formed the group most difficult to cluster in the 1000 Genomes Project data.

@@

>>>>

We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters. 
For this comparison, we use the \ARI{} (ARI) metric, which returns a value between -1 (independent labelings) and 1 (perfect match)~\cite{Hubert1985}.
Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
Including AMR individuals, we observe that the majority of these individuals are placed in the same group as Europeans, likely accurately reflecting their migrational backgrounds. 
Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Hunter2014}.

@@

\subsection*{Personal Genome Project}
To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP). The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features~\cite{Lunshof2013}.
The VCF file we use contains 985,790 variants.

Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
The observed ARI is negative, indicating the clustering is random. 
This poor result may be due to the labels being too broad and not accurately reflecting genetic diversity. For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, and Bulgaria, amongst others. 
The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
% TODO provide mapping
We therefore only include individuals where all grandparents were reported to be from the same country leaving 36 individuals from 23 different countries, which we group into their approximate super-populations (see Supplemental Table X). 
Clustering these individuals results in an ARI of 0.29.

Although we see an increase in cluster-accuracy when removing individuals with a mixed background and operating at super-population level, the clustering is not as precise as on the carefully curated and characterised individuals from the 1000 Genomes Project.
%TODO add numbers
This is especially the case since XX\% are `White' Americans, which formed the group most difficult to cluster in the 1000 Genomes Project data.

@@

\subsection*{ADMIXTURE Implementation}
We use Genome Analysis Toolkit (GATK) to convert the variants from the VCF file, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
These three files are used as input to ADMIXTURE, with K (the number of ancestral populations) set to 4.
